---
title: "Getting started with grin"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with grin}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 6, fig.height = 4.5, dpi = 110
)
```

`grin` turns a 2x2 identification confusion matrix into a calibrated posterior
over the 12 parameters of a General Recognition Theory (GRT) perceptual model.
This vignette walks through a single participant end to end, then a small
sample.

```{r setup}
library(grin)
```

## The ordering contract

A confusion matrix has 4 rows (the true stimuli) and 4 columns (the
responses), and `grin_infer()` needs to know which row/column is which.
**Canonical order** means: number the four stimuli 1-4 by crossing dimension
A's two levels with dimension B's two levels, A changing slower than B:

| position | A level | B level | canonical label |
|:--:|:--:|:--:|:--:|
| 1 | 1 | 1 | `A1B1` |
| 2 | 1 | 2 | `A1B2` |
| 3 | 2 | 1 | `A2B1` |
| 4 | 2 | 2 | `A2B2` |

Row *i* and column *i* of your matrix must be stimulus/response *i* in this
table. Concretely, if dimension A is Age (Old/Young) and dimension B is
Emotion (Neg/Pos): position 1 is Old+Neg, position 2 is Old+Pos, position 3
is Young+Neg, position 4 is Young+Pos -- so row 1 of the matrix holds the
Old+Neg stimulus's responses, in the same four-position order, and so on.

If your matrix is already laid out this way, say so explicitly with
`order = "canonical"` -- `grin_infer()` never assumes it, because a wrong
guess would silently return a confident, wrong posterior:

```{r}
M <- matrix(c(71, 17,  9,  5,
              20, 67,  5,  9,
              13,  6, 63, 20,
               5, 10, 15, 71), nrow = 4, byrow = TRUE)
grin_to_confusion(M, order = "canonical")$counts
```

If it isn't -- your own labelling, a different factor order -- resolve it by
naming the labels you actually have and the two factors they cross, and
`grin_to_confusion()` permutes it into the table above for you:

```{r}
grin_to_confusion(M,
  stim_labels = c("Old/Neg", "Old/Pos", "Young/Neg", "Young/Pos"),
  resp_labels = c("Old/Neg", "Old/Pos", "Young/Neg", "Young/Pos"),
  factor_a = c("Old", "Young"), factor_b = c("Neg", "Pos"))$counts
```

`grin_describe()` prints exactly what will be sent to the network, which is
worth running once on real data before trusting a fit.

## One participant

```{r}
out <- grin_infer(M)
print(out$result)
```

Each of the 12 parameters comes back as an estimate with a posterior SD and a
90% interval: `zx_0`..`zx_3` and `zy_0`..`zy_3` are each stimulus's position
on dimensions A and B (sensitivity, in the identified/standardised space --
0 is chance, ~2.5+ is excellent discrimination), and `rho_0`..`rho_3` are
each stimulus's within-trial perceptual correlation between the two
dimensions (0 = independent, away from 0 = linked or traded off).
`out$result$model_class` is GRIN's best guess at the underlying GRT
structure:

```{r}
out$result$model_class
```

`out$constructs` carries the probabilities behind that label:

```{r}
out$constructs$p_PI
out$constructs$p_sep_A
out$constructs$p_sep_B
```

`evidence_PI`/`evidence_sep_A`/`evidence_sep_B` flag whether this matrix
carries enough information to decide that construct at all -- perceptual
independence in particular is information-limited from a single matrix. A
probability near 0.5 with `evidence_* = FALSE` means "undecided," not
"roughly 50/50 chance."

### Response bias

Separability and independence are about the *shape* of the perceptual
representation. A related but different question is whether the participant
simply favours one response over another, independent of how well they
discriminate -- and GRT, as a multidimensional extension of signal detection
theory, gives two different ways to ask it.

`grin_empirical_bias()` reads it straight off the matrix, no model fit
required: how far each dimension's "respond level 2" rate sits from a fair
coin.

```{r}
grin_empirical_bias(M)
```

`x_bias`/`y_bias` are each in [-0.5, 0.5]: 0 is unbiased, positive means the
participant reports level 2 of that dimension more than a fair coin would,
negative means they favour level 1.

`grin_response_bias()` asks the SDT-native version of the same question: not
how the participant *responded*, but where their *decision criterion* sits.
`grin_infer()`'s identified z-scores place the decision bound at 0 by
convention, so an unbiased observer's two levels on a dimension are mirror
images about it and their four z-scores average to exactly zero -- a nonzero
average is a shifted criterion in the classical SDT sense, read directly off
a fit you already have:

```{r}
grin_response_bias(out$result)
```

Same sign convention as `grin_empirical_bias()` (positive favours level 2),
but a genuinely different quantity: one describes the data, the other
describes the decision rule the fitted model attributes it to, uncertainty
and all (`x_bias_se`/`y_bias_se`, from the fit's own posterior).

## Plotting one participant

```{r, fig.alt = "Perceptual space plot for one participant: four stimulus points in the (zx, zy) plane with correlation ellipses and posterior error bars"}
grin_plot_space(out$result)
```

Four stimulus means in the (zx, zy) plane, one ellipse per stimulus showing
its correlation, dashed lines at the decision bounds (always at 0 in this
identified space). Stimuli are labelled rather than coloured. The ellipse
and the crosshair error bars show two different kinds of uncertainty: the
ellipse is the model's predicted spread of a *single trial's* perceptual
sample (fixed unit variance, shaped by `rho`); the error bars are the
*posterior* uncertainty about where the mean itself sits, given the data.
Turn either off independently (`show_uncertainty = FALSE`), relabel the
stimuli (`stim_labels = ...`), or add per-dimension marginal density strips
(`show_marginals = TRUE`) -- see `?grin_plot_space` for the full set of
arguments (title, axis labels, font size, palette, and more).

```{r, fig.alt = "Forest plot of all 12 parameter estimates with 90 percent credible intervals"}
grin_plot_params(out$result)
```

```{r, fig.alt = "Bar chart of construct probabilities: correlation structure and separability"}
grin_plot_constructs(out$result, out$constructs)
```

Bars for a construct the data can't decide are flagged ("insufficient
evidence") rather than drawn as if informative.

```{r, fig.alt = "Decision-criterion response bias bar chart for dimensions A and B, with error bars"}
grin_plot_bias(out$result)
```

`grin_plot_empirical_bias()` is the matrix-only, no-error-bars counterpart to
this, for the same reason `grin_empirical_bias()` exists alongside
`grin_response_bias()`.

Every plot is a `ggplot` object, so it composes the normal way (more on this
in "Editing a figure further," below).

### Goodness of fit

`grin_plot_diagnostics()` checks the fit against the data it was given,
rather than describing the posterior on its own terms. It needs the original
matrix as well as the fitted result, since `grin_infer()`'s return value
doesn't carry the input back out:

```{r, fig.width = 9, fig.height = 4.5, fig.alt = "Predicted-vs-observed scatter plot and per-dimension marginal density strips"}
grin_plot_diagnostics(out$result, M)
```

The left panel plots the forward model's predicted probability for each of
the 16 stimulus/response cells against its observed proportion -- points on
the diagonal indicate a good fit, and a stimulus that sits off the diagonal
(told apart by point shape, not colour) says where the model is struggling.
The other two panels are the same marginal densities `grin_plot_space()`'s
`show_marginals = TRUE` draws, here paired with the fit check instead. Either
panel can be switched off (`show_predicted_observed = FALSE` or
`show_marginals = FALSE`).

### Palettes

Every plot defaults to black-on-white (`palette = "mono"`) -- publication-
and greyscale-safe, and the only default that doesn't assume anything about
the reader's colour vision. Pass `palette = "<name>"` for one of a few
built-in colour sets, your own character vector of hex colours, or set
`options(grin.palette = "<name>")` once for the session:

```{r}
grin_palette_names()
```

```{r, fig.alt = "The same parameter forest plot rendered with the ember palette"}
grin_plot_params(out$result, palette = "ember")
```

```{r, fig.alt = "The same parameter forest plot rendered with a caller-supplied colour vector"}
grin_plot_params(out$result, palette = c("#123456", "#7A2048", "#1F7A5C"))
```

`"contrast"` is the colour-vision-deficiency-safe categorical palette of
Okabe & Ito (2008); `grin_plot_space()` and `grin_plot_space_group()` never
split one participant's four stimuli by colour regardless of palette --
stimuli are told apart by their label, since with everything else on the
plot fixed a fifth encoding channel would be redundant with it.

## Stopping rules for adaptive designs

If you're collecting data adaptively -- more trials until you know enough --
declare what "enough" means as a `grin_criterion()` of targets, then check it
after each `grin_infer()`:

```{r}
crit <- grin_criterion(list(
  grin_target_precision(params = c("zx", "zy"), sd_max = 0.10),   # want the space measured
  grin_target_probability("PS_A", at_least = 0.90)                # want the verdict
), combine = "any")

decision <- grin_evaluate(crit, out$result, out$constructs)
print(decision)
```

Here the precision target isn't met yet (`zx_3`'s SD is still above 0.10),
but the probability target is, so with `combine = "any"` the loop stops. A
threshold on a construct the data cannot decide is reported in
`decision$blocked_by` and never stops the loop.

## Many participants

Loop `grin_infer()` over a sample and pass the (optionally named) list
straight to the group-level tools:

```{r}
mats <- list(
  p1 = matrix(c(71, 17,  9,  5, 20, 67,  5,  9,
                13,  6, 63, 20,  5, 10, 15, 71), nrow = 4, byrow = TRUE),
  p2 = matrix(c(50, 10, 15, 25, 12, 55, 20, 13,
                18, 22, 48, 12,  8, 14, 18, 60), nrow = 4, byrow = TRUE),
  p3 = matrix(c(65, 20, 10,  5, 15, 70,  3, 12,
                 8,  4, 68, 20,  3, 10, 18, 69), nrow = 4, byrow = TRUE),
  p4 = matrix(c(45, 40, 10,  5, 38, 47,  7,  8,
                 6,  9, 48, 37,  8,  7, 42, 43), nrow = 4, byrow = TRUE),
  p5 = matrix(c(30, 28, 22, 20, 26, 29, 24, 21,
                22, 25, 27, 26, 19, 23, 26, 32), nrow = 4, byrow = TRUE),
  p6 = matrix(c(88,  5,  5,  2,  4, 90,  2,  4,
                 5,  3, 87,  5,  2,  4,  6, 88), nrow = 4, byrow = TRUE)
)
sample <- lapply(mats, grin_infer)
```

`grin_tidy()` is the shared foundation every group plot builds on:

```{r}
td <- grin_tidy(sample)
td[, c("id", "model_class", "p_PI", "p_sep_A", "p_sep_B")]
```

```{r, fig.width = 8, fig.height = 5, fig.alt = "One perceptual-space panel per participant, faceted"}
grin_plot_space_group(sample)
```

One panel per participant (`facet = TRUE`, the default) so individual
uncertainty stays visible -- this is the individual-level analysis GRT
actually licenses.

`facet = FALSE` overlays everyone instead, with a single across-participant
mean ellipse per stimulus. **This is an exploratory inspection view only,
not a reporting figure**: GRT's perceptual space is defined per observer,
and a "grand mean" ellipse over several independently-fitted spaces is not
itself a fitted GRT model. Use it to eyeball whether a sample looks roughly
homogeneous before deciding how to report it properly -- not as the figure
itself. `grin` prints a reminder every time this mode runs:

```{r, fig.alt = "All participants' stimulus points overlaid with a single across-participant mean ellipse"}
grin_plot_space_group(sample, facet = FALSE)
```

```{r, fig.width = 8, fig.alt = "Boxplots of each of the 12 parameters' estimates across participants"}
grin_plot_params_group(sample)
```

```{r, fig.alt = "Bar chart of how many participants fall in each inferred GRT model class"}
grin_plot_model_classes(sample)
```

```{r, fig.alt = "Boxplots of posterior SD per parameter group across participants -- data quality across the sample"}
grin_plot_precision_group(sample)
```

```{r, fig.alt = "Boxplots of decision-criterion response bias per dimension across participants"}
grin_plot_bias_group(sample)
```

`grin_plot_empirical_bias_group(mats)` is the matrix-only counterpart, taking
the raw confusion matrices directly rather than fitted results.

`grin_plot_precision_group()` is useful for planning: run it on pilot data to
see what posterior SD your trial counts are actually buying you, before
picking a `sd_max` for `grin_target_precision()`.

## Editing a figure further

Every `grin_plot_*()` returns a plain `ggplot` object (or, for the
multi-panel functions, a `patchwork` object built from plain `ggplot`
objects), so it takes ordinary ggplot2 calls rather than needing a package
argument for every possible customisation:

```{r, fig.alt = "The space plot with a manually replaced title and a manually added reference annotation"}
library(ggplot2)

grin_plot_space(out$result) +
  labs(title = "Participant 7, session 2", subtitle = NULL) +
  annotate("text", x = 0, y = 2.3, label = "note: post-training session",
          size = 3, colour = "grey40")
```

Anything ggplot2 can do to a plot -- swapping the theme, changing a scale,
faceting further, saving at a specific size with `ggsave()` -- works the
normal way here too. That's deliberate: this package covers the arguments
that come up often (title, labels, palette, uncertainty, marginals); for
everything else, editing the returned object is the intended path, not a
workaround.

## Where next

- `?grin_infer`, `?grin_to_confusion`, `?grin_criterion`,
  `?grin_response_bias` -- full argument reference for everything used above.
- The package [README](https://github.com/MurraySBennett/grin/tree/main/packages/grin)
  for the Python (`grintools`) equivalent of every function here.
- [Interpreting GRIN output](https://github.com/MurraySBennett/grin/blob/main/docs/interpreting.md)
  -- a plain-language guide to reading the numbers above.
- [The GRT model specification](https://github.com/MurraySBennett/grin/blob/main/docs/GRT_model_spec.md)
  -- the underlying math: the 12 parameters, the model classes, the sign
  convention, decisional separability.
