---
title: "Getting started with grin"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with grin}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 6, fig.height = 4.5, dpi = 110
)
```

`grin` turns a 2x2 identification confusion matrix into a calibrated posterior
over the 12 parameters of a General Recognition Theory (GRT) perceptual model
-- in milliseconds, via a neural network trained once and shipped inside the
package, rather than a per-participant maximum-likelihood fit. This vignette
walks through a single participant end to end, then a small sample.

```{r setup}
library(grin)
```

## One participant

A confusion matrix in **canonical order** -- rows and columns
`A1B1, A1B2, A2B1, A2B2` (dimension A varies slower than B):

```{r}
M <- matrix(c(71, 17,  9,  5,
              20, 67,  5,  9,
              13,  6, 63, 20,
               5, 10, 15, 71), nrow = 4, byrow = TRUE)

out <- grin_infer(M)
print(out$result)
```

If your data isn't already in this order -- different stimulus/response
labelling, factors in a different sequence -- resolve it explicitly with
`grin_to_confusion()` rather than guessing; see `?grin_to_confusion` and the
README. A silent row/column swap would return a confident wrong posterior, so
`grin_infer()` trusts you've already done this.

Each of the 12 parameters comes back as an estimate with a posterior SD and a
90% interval: `zx_0`..`zx_3` and `zy_0`..`zy_3` are each stimulus's position
on dimensions A and B (sensitivity, in the identified/standardised space --
0 is chance, ~2.5+ is excellent discrimination), and `rho_0`..`rho_3` are each
stimulus's within-trial perceptual correlation between the two dimensions
(0 = independent, away from 0 = linked or traded off). `out$result$model_class`
is GRIN's best guess at the underlying GRT structure:

```{r}
out$result$model_class
```

`out$constructs` carries the probabilities behind that label -- how likely the
correlation structure is perceptual independence (PI) / one shared correlation
(RHO1) / a free correlation per stimulus, and how likely each dimension is
perceptually separable:

```{r}
out$constructs$p_PI
out$constructs$p_sep_A
out$constructs$p_sep_B
```

`evidence_PI`/`evidence_sep_A`/`evidence_sep_B` flag whether this matrix
actually carries enough information to decide that construct at all --
perceptual independence in particular is information-limited from a single
matrix. A probability near 0.5 with `evidence_* = FALSE` means "undecided,"
not "roughly 50/50 chance."

## Plotting one participant

```{r, fig.alt = "Perceptual space plot for one participant: four stimulus points in the (zx, zy) plane with correlation ellipses"}
grin_plot_space(out$result)
```

Four stimulus means in the (zx, zy) plane, one ellipse per stimulus showing
its correlation, dashed lines at the decision bounds (always at 0 in this
identified space -- that's the point of working in it). Stimuli are told
apart by a label, not a colour: with everything else about the plot fixed,
four colours for four points is decoration, not information. See "Style"
below for switching this on anyway.

```{r, fig.alt = "Forest plot of all 12 parameter estimates with 90 percent credible intervals"}
grin_plot_params(out$result)
```

```{r, fig.alt = "Bar chart of construct probabilities: correlation structure and separability"}
grin_plot_constructs(out$result, out$constructs)
```

Bars for a construct the data can't decide are visually flagged
("insufficient evidence") rather than drawn as if they were informative --
same principle as `evidence_*` above.

Every plot is a `ggplot` object, so it composes the normal way:
`grin_plot_space(out$result) + ggplot2::labs(title = "Participant 7")`.

### Style

Every `grin_plot_*()` defaults to black-on-white (publication- and
greyscale-safe). Pass `color = TRUE` for the house blue/rose style instead,
or set it once for the session with `options(grin.color = TRUE)`; an
explicit `color =` argument always overrides the option.

```{r, fig.alt = "The same parameter forest plot in the optional colour style"}
grin_plot_params(out$result, color = TRUE)
```

## Stopping rules for adaptive designs

If you're collecting data adaptively -- more trials until you know enough --
declare what "enough" means as a `grin_criterion()` of targets, then check it
after each `grin_infer()`:

```{r}
crit <- grin_criterion(list(
  grin_target_precision(params = c("zx", "zy"), sd_max = 0.10),   # want the space measured
  grin_target_probability("PS_A", at_least = 0.90)                # want the verdict
), combine = "any")

decision <- grin_evaluate(crit, out$result, out$constructs)
print(decision)
```

Here the precision target isn't met yet (`zx_3`'s SD is still above 0.10),
but the probability target is, so with `combine = "any"` the loop stops. A
threshold on a construct the data cannot decide (an `evidence_* = FALSE`
construct) is reported in `decision$blocked_by` and never stops the loop --
the limit is in the data, not the tool.

## Many participants

Loop `grin_infer()` over a sample and pass the (optionally named) list
straight to the group-level tools:

```{r}
mats <- list(
  p1 = matrix(c(71, 17,  9,  5, 20, 67,  5,  9,
                13,  6, 63, 20,  5, 10, 15, 71), nrow = 4, byrow = TRUE),
  p2 = matrix(c(50, 10, 15, 25, 12, 55, 20, 13,
                18, 22, 48, 12,  8, 14, 18, 60), nrow = 4, byrow = TRUE),
  p3 = matrix(c(65, 20, 10,  5, 15, 70,  3, 12,
                 8,  4, 68, 20,  3, 10, 18, 69), nrow = 4, byrow = TRUE),
  p4 = matrix(c(45, 40, 10,  5, 38, 47,  7,  8,
                 6,  9, 48, 37,  8,  7, 42, 43), nrow = 4, byrow = TRUE),
  p5 = matrix(c(30, 28, 22, 20, 26, 29, 24, 21,
                22, 25, 27, 26, 19, 23, 26, 32), nrow = 4, byrow = TRUE),
  p6 = matrix(c(88,  5,  5,  2,  4, 90,  2,  4,
                 5,  3, 87,  5,  2,  4,  6, 88), nrow = 4, byrow = TRUE)
)
sample <- lapply(mats, grin_infer)
```

`grin_tidy()` is the shared foundation every group plot builds on -- one row
per participant, in case you want the raw numbers instead of a plot:

```{r}
td <- grin_tidy(sample)
td[, c("id", "model_class", "p_PI", "p_sep_A", "p_sep_B")]
```

This toy sample is illustrative, not simulated from known ground truth, but
it's a real, if small, spread: two participants (`p2`, `p6`) come back with a
`free`-correlation structure rather than PI, and precision (the width of
each interval) varies with how the counts happen to fall. That variation is
exactly what the group plots below are for.

```{r, fig.width = 8, fig.height = 5, fig.alt = "One perceptual-space panel per participant, faceted"}
grin_plot_space_group(sample)
```

One panel per participant (`facet = TRUE`, the default) so individual
uncertainty stays visible. Set `facet = FALSE` to overlay everyone instead,
with faint individual points and a labelled group-mean ellipse per stimulus
-- readable with more participants than a facet grid can hold, at the cost
of hiding each person's own uncertainty:

```{r, fig.alt = "All participants' stimulus points overlaid with group-mean ellipses"}
grin_plot_space_group(sample, facet = FALSE)
```

```{r, fig.width = 8, fig.alt = "Boxplots of each of the 12 parameters' estimates across participants"}
grin_plot_params_group(sample)
```

```{r, fig.alt = "Bar chart of how many participants fall in each inferred GRT model class"}
grin_plot_model_classes(sample)
```

```{r, fig.alt = "Boxplots of posterior SD per parameter group across participants -- data quality across the sample"}
grin_plot_precision_group(sample)
```

`grin_plot_precision_group()` is useful for planning: run it on pilot data to
see what posterior SD your trial counts are actually buying you, before
picking a `sd_max` for `grin_target_precision()`.

## Where next

- `?grin_infer`, `?grin_to_confusion`, `?grin_criterion` -- full argument
  reference for everything used above.
- The package [README](https://github.com/MurraySBennett/grin/tree/main/packages/grin)
  for the Python (`grintools`) equivalent of every function here, and the
  differences in API convention between the two.
- [Interpreting GRIN output](https://github.com/MurraySBennett/grin/blob/main/docs/interpreting.md)
  -- a plain-language guide to reading the numbers above.
- [The GRT model specification](https://github.com/MurraySBennett/grin/blob/main/docs/GRT_model_spec.md)
  -- the underlying math: the 12 parameters, the model classes, the sign
  convention, decisional separability.
